Make any stupid plot you want

### 2021-04-28 draft/sections/intro.tex
new intro

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Visualizations are made up of transformations from data into visual representation. These functions transform individual data values to visual representation, such as date to x position or latitude to y position. These functions are composed into the assembly of all these transformations into a visual mark, such as a line or point. %%% break this out into seperate sentences to break 
    The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
    \label{fig:intro:artist}
\end{figure}
Visualization is the transformation of data into visual representation. As illustrated by \autoref{fig:intro:artist}, these translations are both at the level of the individual variable and the entire record. In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors. A row in the table is collectively encoded as a point mark. None of these encodings are fixed, as evidenced by temperature being translated into the y value in the case of the line plot. The station is now the source of the color of the entire line, and the date is the x position. As with scatter, the encodings of the individual transformations, which again are on values from the same record in the table, are composited into a line mark. It is these raw transformations from data space to visualization space that are implemented by building block level visualization libraries, named as such because the functions provided by the library can be composited in any number of ways to yield visualizations \cite{wongsuphasawatNavigatingWideWorld2021}. We propose that like physical building blocks, building block libraries must provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together. We specify that a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}. We then use this model to develop a design specification for the components of a building block visualization library. The notion of self contained, inherently modular, building blocks lends itself naturally to a functional paradigm of visualization \cite{hughesWhyFunctionalProgramming1989}. We adopt a functional model for a redesign because the lack of side effects means functional architecture can be evaluated for correctness, functional programs tend to be shorter and clearer, and are well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library. One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner. Particularly, each new component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is then expressed in the API for that component. In turn, this yields an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}. This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g. shape or color, are supported. We propose that these issues can be ameliorated via a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints. We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints. We then use the common data model introduced by the model to demonstrate how plotting functions can be consolidated in a way that makes clear whether the difference is in expected data structure, visual component encoding, or the resulting graphic.

@@

\subsection{Structure:}
\label{sec:intro:structure}
Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Building block visualization libraries implement the functions that encode data as visual representations. For example, functions that map latitude and longitude into positions on a graphical representation of a map and functions that map precipitation and station name to colors. Visualization libraries also provide functions that composite these encoding functions into attributes of a visual element, such as a point, and we call these functions \textit{artists}.  The encoders and \textit{artists} are reusable, so position encoders can also be used to to translate date and time to positions, which can then be composited into a line by a \texttt{line} function. 
    }
    \label{fig:intro:artist}
\end{figure}

Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. We define visualization components to be structure preserving maps from data to visual representations. As illustrated in \autoref{fig:intro:artist}, visualizations map data into graphical elements such as points and lines and map individual data components (columns) into components of the visualization such as position or color. For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors. A \texttt{point} function composites these encodings into attributes of a point on screen. The \texttt{line} function composites position and color encodings into a line, specifically encoders that convert date and precipitation to x and y positions and a map from station name to color. We identify the structure these maps must preserved as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.

We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification. Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.  Just as physical building blocks can be arranged in an almost infinite number of ways, solely constrained by the size and shape of the blocks, we propose that the only restriction on the composition of building block library components are \textit{continuity} and \textit{equivariance}. We adopt a functional model for our proposed architecture because independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}. Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library. We even name these compositing functions \texttt{point} and \texttt{line} in \autoref{fig:intro:artist} \textit{artists} as a nod to the somewhat analogous object in the Matplotlib architecture.  One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner. Particularly, each new component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is then expressed in the API for that component. In turn, this yields an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}. This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g. shape or color, are supported. We propose that these issues can be ameliorated via a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints. We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints. We then use the common data model introduced by the model to demonstrate how plotting functions can be consolidated in a way that makes clear whether the difference is in expected data structure, visual component encoding, or the resulting graphic.

@@

\subsection{Structure:}
\label{sec:intro:structure}
 The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
